Wrangling multivariate spatio-temporal data with the R package cubble

### 2022-04-28 1-intro.Rmd
address patricia's comment in intro

The SpatioTemporal CRAN task view [@ctvspatiotemporal] gathers information about R packages designed for spatio-temporal data and it has a section on *Representing data* that lists existing spatio-temporal data representations used in \proglang{R}. Among them, @spacetime proposes four spatio-temporal layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} [@spacetime] package via the underlying spatial and temporal classes \pkg{sp} [@sp] and \pkg{xts} [@xts]. The \pkg{stars} [@stars] package has been a recent implementation to handle both raster and vector data using spatio-temporal arrays. It interfaces with \pkg{sf} [@sf], which adopts the simple features representation of spatial data, and \pkg{tidyverse} [@tidyverse], a suite of tools to wrangle and visualise data in \proglang{R}.

**The paragraph above needs a bit more flesh**

Still, the data representation for spatio-temporal data can be further improved and there are two reasons for this. Firstly, the raw data sourced in the wild is less often presented in any one of the representations above (Which representations - in the previous paragraph there are more than two?), and fitting the raw data into a data object can sometimes be difficult. More often, spatio-temporal data are collected in separate 2D tables and analysts need to assemble them into a whole piece before exploring the data. Examples of components of spatio-temporal data can be 1) map data recording the shape of a collection of areas of interest; 2) geo-scientific data recording the longitude and latitude coordinates of locations in the area along with other metadata related to the location, and; 3) temporal variables indexed by both location and time.

**I would aligned the examples of spatio-temporal data with those discuss in the st books. See Paula Moraga for example**

@@

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately or combined while being kept synchronised. Among the four spacetime layouts in @spacetime, \pkg{cubble} can be applied to full di layout, sparse grid layout, or irregular layout, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at https://CRAN.R-project.org/package=cubble. **At this point what is cubble is adding with respect the other packages it is not  clear to me**

>>>>

The SpatioTemporal CRAN task view [@ctvspatiotemporal] gathers information about R packages designed for spatio-temporal data and it has a section on *Representing data* that lists existing spatio-temporal data representations used in \proglang{R}. Among them, @spacetime summarises spatio-temporal data into three forms: time-wide, space-wide, and long formats. The associated package \pkg{spacetime} [@spacetime] implements four spatio-temporal layouts (full grid, sparse grid, irregular, and trajectory) to handle different space and time combinations. The \pkg{stars} [@stars] package has been a recent implementation to use dense arrays to represent spatio-temporal cubes. It also interfaces with \pkg{sf} [@sf], a package commonly used for wrangling spatial data, and the \pkg{tidyverse} [@tidyverse] suite for general data wrangling and visualisation in \proglang{R}.

Still, the data representation for spatio-temporal data can be further improved and there are two reasons for this. Firstly, the raw data sourced in the wild is less often presented in any one of the layouts above, and fitting the raw data into a data object can sometimes be difficult. More often, spatio-temporal data are collected in separate 2D tables and analysts need to assemble them into a whole piece before exploring the data. Examples of components of spatio-temporal data can be 1) areal data recording the shape of a collection of areas of interest; 2) geostatistical data recording the longitude and latitude coordinates of locations, potentially also with other metadata related to the location, and; 3) temporal data of each location across time.

<!-- **I would aligned the examples of spatio-temporal data with those discuss in the st books. See Paula Moraga for example** -->

@@

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately or combined while being kept synchronised. Among the four spacetime layouts in @spacetime, \pkg{cubble} can be applied to full grid, sparse grid, or irregular, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at https://CRAN.R-project.org/package=cubble. **At this point what is cubble is adding with respect the other packages it is not  clear to me**

### 2022-04-28 arxiv/cubble-arxiv.tex
address patricia's comment in intro

The SpatioTemporal CRAN task view (\protect\hyperlink{ref-ctvspatiotemporal}{Edzer Pebesma 2022}) gathers information about R packages designed for spatio-temporal data and it has a section on \emph{Representing data} that lists existing spatio-temporal data representations used in \proglang{R}. Among them, E. Pebesma (\protect\hyperlink{ref-spacetime}{2012}) proposes four spatio-temporal layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} (\protect\hyperlink{ref-spacetime}{E. Pebesma 2012}) package via the underlying spatial and temporal classes \pkg{sp} (\protect\hyperlink{ref-sp}{E. Pebesma and Bivand 2005}) and \pkg{xts} (\protect\hyperlink{ref-xts}{Ryan and Ulrich 2020}). The \pkg{stars} (\protect\hyperlink{ref-stars}{E. Pebesma 2021}) package has been a recent implementation to handle both raster and vector data using spatio-temporal arrays. It interfaces with \pkg{sf} (\protect\hyperlink{ref-sf}{E. J. Pebesma 2018}), which adopts the simple features representation of spatial data, and \pkg{tidyverse} (\protect\hyperlink{ref-tidyverse}{Wickham et al. 2019}), a suite of tools to wrangle and visualise data in \proglang{R}.

\textbf{The paragraph above needs a bit more flesh}

Still, the data representation for spatio-temporal data can be further improved and there are two reasons for this. Firstly, the raw data sourced in the wild is less often presented in any one of the representations above (Which representations - in the previous paragraph there are more than two?), and fitting the raw data into a data object can sometimes be difficult. More often, spatio-temporal data are collected in separate 2D tables and analysts need to assemble them into a whole piece before exploring the data. Examples of components of spatio-temporal data can be 1) map data recording the shape of a collection of areas of interest; 2) geo-scientific data recording the longitude and latitude coordinates of locations in the area along with other metadata related to the location, and; 3) temporal variables indexed by both location and time.

\textbf{I would aligned the examples of spatio-temporal data with those discuss in the st books. See Paula Moraga for example}

@@

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately or combined while being kept synchronised. Among the four spacetime layouts in E. Pebesma (\protect\hyperlink{ref-spacetime}{2012}), \pkg{cubble} can be applied to full di layout, sparse grid layout, or irregular layout, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at \url{https://CRAN.R-project.org/package=cubble}. \textbf{At this point what is cubble is adding with respect the other packages it is not clear to me}

@@

\leavevmode\vadjust pre{\hypertarget{ref-sp}{}}%
Pebesma, Edzer, and Roger S Bivand. 2005. {``S Classes and Methods for Spatial Data: The \pkg{sp} Package.''} \emph{R News} 5 (2): 9--13.

@@

\leavevmode\vadjust pre{\hypertarget{ref-xts}{}}%
Ryan, Jeffrey A., and Joshua M. Ulrich. 2020. \emph{\pkg{xts}: eXtensible Time Series}. \url{https://CRAN.R-project.org/package=xts}.

>>>>

The SpatioTemporal CRAN task view (\protect\hyperlink{ref-ctvspatiotemporal}{Edzer Pebesma 2022}) gathers information about R packages designed for spatio-temporal data and it has a section on \emph{Representing data} that lists existing spatio-temporal data representations used in \proglang{R}. Among them, E. Pebesma (\protect\hyperlink{ref-spacetime}{2012}) summarises spatio-temporal data into three forms: time-wide, space-wide, and long formats. The associated package \pkg{spacetime} (\protect\hyperlink{ref-spacetime}{E. Pebesma 2012}) implements four spatio-temporal layouts (full grid, sparse grid, irregular, and trajectory) to handle different space and time combinations. The \pkg{stars} (\protect\hyperlink{ref-stars}{E. Pebesma 2021}) package has been a recent implementation to use dense arrays to represent spatio-temporal cubes. It also interfaces with \pkg{sf} (\protect\hyperlink{ref-sf}{E. J. Pebesma 2018}), a package commonly used for wrangling spatial data, and the \pkg{tidyverse} (\protect\hyperlink{ref-tidyverse}{Wickham et al. 2019}) suite for general data wrangling and visualisation in \proglang{R}.

Still, the data representation for spatio-temporal data can be further improved and there are two reasons for this. Firstly, the raw data sourced in the wild is less often presented in any one of the layouts above, and fitting the raw data into a data object can sometimes be difficult. More often, spatio-temporal data are collected in separate 2D tables and analysts need to assemble them into a whole piece before exploring the data. Examples of components of spatio-temporal data can be 1) areal data recording the shape of a collection of areas of interest; 2) geostatistical data recording the longitude and latitude coordinates of locations, potentially also with other metadata related to the location, and; 3) temporal data of each location across time.

@@

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately or combined while being kept synchronised. Among the four spacetime layouts in E. Pebesma (\protect\hyperlink{ref-spacetime}{2012}), \pkg{cubble} can be applied to full grid, sparse grid, or irregular, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at \url{https://CRAN.R-project.org/package=cubble}. \textbf{At this point what is cubble is adding with respect the other packages it is not clear to me}

@@

@@
